Discovering cell type motifs in connectomics data

### 2013-12-04 docs/paper.tex
updates to writing; addition of todos

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{subfiles}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{multicol}

@@

% \begin{affiliations}
%   \item Something Something
%   \item Something Something
% \end{affiliations}

@@

\section*{Introduction}
Emerging techniques (cite Lichtman, Zador, Denk etc) promise to
quantify the location of each neuron within a volume of interest,
along with its connections to all other neurons. Far exceeding the
capacity of neuroanatomists to trace small circuits, this leads to
enormous datasets that quantify aspects of the nervous system. The
rise of high-throughput sequencing techniques necessitated the
development of novel computational methods to understand genomic
structure, ushering in an era of bioinformatics as an independent
discipline (cite).

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity motifs
has been prominent in many areas, ranging from sensory (e.g. retina,
cite) via processing (e.g. neocortex) to movement (e.g. spinal cord)
(cites). Indeed, such motifs also exists in human-made computing
circuits. It remains an important challenge to develop algorithms to
use anatomical data, e.g. connectomics, to back out underlying
connectivity motifs.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network (cite Strogatz scale-free
stuff).  While this approach leads to an understanding of global
network, more recent work aims at identifying repeat patterns, or
“motifs” in networks. A motif is a set of nodes with conserved
relations that appears multiple times in a network, for example a set
of three nodes that are each connected to one another. When it comes
to neuroscience, we might expect that a motif consists of the kinds of
cells that jointly make up a microcircuit along with their distance
and cell-type dependent connectivity. The same motif should then
appear in multiple locations in the nervous system. Combining
motif-discovery ideas with specific knowledge about the kinds of
motifs in the nervous system promises better structure discovery.

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include spectral methods
(cite), community-based-detection methods (cite), and stochastic block
models (cite). While these approaches powerfully incorporate the
probabilistic nature of neural connections (cite) they do not
generally consider the known fact that the probability of connections
between neurons depends on distance. Combining such probabilistic
models with a spatial component to model distance promises to improve
the detection of microcircuitry from neural data.

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: both the cell
types and their spatial patterns of interconnection. We apply it to
three computing systems: the mouse retina connectome (cite), the
c. elegans connectome (cite), and a ``connectome'' of a classical
microprocessor. In all cases, we discover cell-type to cell-type
connectivity that depends on spatial distance and these connectivity
functions depend on the two involved cell types. Comparing the cell
types discovered by the algorithms with those obtained by anatomists
based on orthogonal information reveals a high degree of agreement. We
present a scalable approach to infer microcircuitry from connectivity
data.

@@

\subsubsection{Mouse Retina}
The retina of the mouse (cite) is an example where we should expect to
have connectivity patterns that are well approximated by our
generative model. It is known that there are multiple classes of cells
that can be grouped into ganglion cells that transmit information to
the rest of the brain, bipolar cells that connect between different
cells, and amacrine cells that feed into the ganglion cells (cite some
review). Recent research (cite) has produced a large dataset
containing both the types of cells from orthogonal approaches, and
also the connectivity matrix between all reconstructed cells. We want
to know to which level we can classify neurons into types based on
this connectivity only.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience, with the location and connectivity of each
of 302 neurons deterministically known. This incredible cross-organism
consistency led to early mapping of the c. elegans chemical and
electrical (gap junction) connectome. Unlike the retina, only the
motor neurons in c. elegans exhibit regular distribution in space, as
they follow the body axis. Most interneurons are concentrated in
various ganglia that project throughout the entire animal, and the
sensory neurons are primarily located in the animal head.

@@

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of a
classic microprocessor, the MOS Technology 6502 used in the original
Apple II, Atari, and Commodore 64 (cite datasource). We identified a
region of the processor with complex but known structure containing
the primary general-purpose registers X, Y, and S (figure).

@@

For probabilistic models like ours, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns [homogenize these two wordings across the
paper]. We employ a collection of Markov-chain Monte carlo techniques
[reference methods] but while different initializations converge to
similar ultimate values, we can never realistically obtain the global
optimum. There are a broad range of techniques that may offer better
approximations to the global optimum (cite) and future work could
adapt them to the problem we addressed here.

Our inference becomes slower as the amount of data increases. Our
algorithm required 10xxx hours for 1000 neurons and would therefore
take months for 10,000 neurons, which would not be acceptable. Various
new techniques promise to allow similar calculations to take less time
and scale to larger datasets (cites).

@@

We assume as input a collection of connectivity matrices between
cells, as well as a distance function $d(e_i, e_j)$. We assume there
exist an unknown number of latent (unobserved) cell types, $k$, $k \in
\{1, 2, 3, \dots\}$, and that each cell $e_i$ belongs to a single cell
type, $c_i = k$. The observed connectivity between two cells $R(e_i,
e_j)$ then only depends on their latent type and their distance
through a link function $f(c_i, c_j, d(e_i, e_j))$ If $c_i=m$ and
$c_j=n$ then there exists a latent-class-specific set of parameters
$\eta_{mn}$ which parametrizes $f$, as well as a set of global hyper parameters $\theta$.

@@

\begin{equation}
  \P(\vec{c}, \eta, \theta) \propto \prod_{i, j} P() \prod_{m, n} P(\eta_{mn} | \theta)  P(\theta) P(c | \alpha) p(\alpha) 
\end{equation}

@@

Our model is a extension of infinite stochastic block models
\cite{Kemp, other guys} to incorporate spatial constraints.

@@

The ``logistic-distance Bernoulli'' model associates with each latent class 
two parameters, $\mu_{mn}$ and $\lambda_{mn}$, and the probability that two
cells $e_i$ and $e_j$ are connected is given by
\begin{equation}
p^* = \frac{1.0}{1 + \exp \frac{d(e_i, e_j) - \mu_{mn}}{\lambda_{mn}}}\\
p = p^* \cdot (p_{max} - p_{min}) + p_{min}
\end{equation}
where $p_{max}$ and $p_{min}$ are per-graph hyperparameters. Per-component parameters $\mu_{mn} \sim \exp(\mu | \mu^{hp})$ and $\lambda_{mn} \sim \exp(\lambda | \lambda^{hp})$. 
Inference is performed on these hyperparameters as well, see TODO supplemental 
for details.

@@

We then employ a transition kernel to update the per-component
parameter values $\eta_{mn}$. Conditioned on the assignment vector
$\vec{c}$ and the model hyperparameters $\theta, \alpha$ the 
individual $\eta_{mn}$ are indepednent. We slice sample \autocite{NealSlice}
each component's parameters, choosing the slice width as a function
of the global hyperparameter range.

@@

We obtained the connectome of c. elegans from \autocite, and isolated
the 279 nonpharyngeal neurons, with a total of 6393 chemical synapses
and 890 gap junctions originally cleaned up in \autocite{Chen2006}. A
cell's position was its distance along the anterior-posterior axis
normalzed between zero and one. We used both networks, the chemical
network as a directed graph and the electrical network as undirected
graph. We use the synapse counts with the logistic-distance poisson
likelihood, scaling the counts by a factor of 4 to compensate for 
the Poisson's underdispersion.

% \begin{addendum}
%  \item Put acknowledgements here.
%  \item[Competing Interests] The authors declare that they have no
% competing financial interests.
%  \item[Correspondence] Correspondence and requests for materials
% should be addressed to A.B.C.~(email: myaddress@nowhere.edu).
% \end{addendum}

>>>>

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{subfiles}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{multicol}

@@

\todo{Citations}
\todo{Log score plots for real data} 
\todo{Fix vector notation}
\todo{Get feedback on joint model score}
\todo{Find other irm paper citation}
% \begin{affiliations}
%   \item Something Something
%   \item Something Something
% \end{affiliations}

@@

\section*{Introduction}
Emerging techniques \autocite{Lichtman, Zador, Denk} promise to
quantify the location of each neuron within a volume of interest,
along with its connections to all other neurons. Far exceeding the
capacity of neuroanatomists to trace small circuits, this leads to
enormous datasets that quantify aspects of the nervous system. The
rise of high-throughput sequencing techniques necessitated the
development of novel computational methods to understand genomic
structure, ushering in an era of bioinformatics as an independent
discipline \autocite{}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity motifs
has been prominent in many areas, ranging from sensory (e.g. retina,
\autocite{} via processing (e.g. neocortex) to movement (e.g. spinal cord)
\autocite{}. Indeed, such motifs also exists in human-made computing
circuits. It remains an important challenge to develop algorithms to
use anatomical data, e.g. connectomics, to back out underlying
connectivity motifs.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network \autocite {WattsStrogatz1998}
.  While this approach leads to an understanding of global
network, more recent work aims at identifying repeat patterns, or
“motifs” in networks. \todo{There's the Science2002 paper by Milo et. al. that explicitly looks at 'motifs'} A motif is a set of nodes with conserved
relations that appears multiple times in a network, for example a set
of three nodes that are each connected to one another. When it comes
to neuroscience, we might expect that a motif consists of the kinds of
cells that jointly make up a microcircuit along with their distance
and cell-type dependent connectivity. The same motif should then
appear in multiple locations in the nervous system. Combining
motif-discovery ideas with specific knowledge about the kinds of
motifs in the nervous system promises better structure discovery.

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include spectral methods
\autocite{}, community-based-detection methods \autocite{}, and stochastic block
models \autocite{NowickiSnijdders2001}. While these approaches powerfully incorporate the
probabilistic nature of neural connections \autocite{} they do not
generally consider the known fact that the probability of connections
between neurons depends on distance. Combining such probabilistic
models with a spatial component to model distance promises to improve
the detection of microcircuitry from neural data.

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: both the cell
types and their spatial patterns of interconnection. We apply it to
three computing systems: the mouse retina connectome \autocite{}, the
c. elegans connectome \autocite{}, and a ``connectome'' of a classical
microprocessor. In all cases, we discover cell-type to cell-type
connectivity that depends on spatial distance and these connectivity
functions depend on the two involved cell types. Comparing the cell
types discovered by the algorithms with those obtained by anatomists
based on orthogonal information reveals a high degree of agreement. We
present a scalable approach to infer microcircuitry from connectivity
data.

@@

\subsubsection{Mouse Retina}
The retina of the mouse \autocite{Masland2001} is an example where we
should expect to have connectivity patterns that are well approximated
by our generative model. It is known that there are multiple classes
of cells that can be grouped into ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells (cite some review). Recent research \autocite{Helmstaedter2013} has
produced a large dataset containing both the types of cells from
orthogonal approaches, and also the connectivity matrix between all
reconstructed cells. We want to know to which level we can classify
neurons into types based on this connectivity only.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, as they follow the body axis. Most interneurons
are concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in the animal
head.

@@

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of a
classic microprocessor, the MOS Technology 6502 used in the original
Apple II, Atari, and Commodore 64 \autocite{datasource}. We identified a
region of the processor with complex but known structure containing
the primary general-purpose registers X, Y, and S (figure).

@@

For probabilistic models like ours, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns [homogenize these two wordings across the
paper]. We employ a collection of Markov-chain Monte carlo techniques
[reference methods] but while different initializations converge to
similar ultimate values, we can never realistically obtain the global
optimum. There are a broad range of techniques that may offer better
approximations to the global optimum \autocite{Wanga2012,FritzJonathan} and future work could
adapt them to the problem we addressed here.

Our inference becomes slower as the amount of data increases. Our
algorithm required 10xxx hours for 1000 neurons and would therefore
take months for 10,000 neurons, which would not be acceptable. Various
new techniques promise to allow similar calculations to take less time
and scale to larger datasets \autocite{}.

@@

We assume as input a connectivity matrix $R$ defining the connections
between cells, as well as a distance function $d(e_i, e_j)$
representing a (physical) distance between adjacent cells. See the
supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number of latent
(unobserved) cell types, $k$, $k \in \{1, 2, 3, \dots\}$, and that
each cell $e_i$ belongs to a single cell type, $c_i = k$. The observed
connectivity between two cells $R(e_i, e_j)$ then only depends on
their latent type and their distance through a link function $f(c_i,
c_j, d(e_i, e_j))$ If $c_i=m$ and $c_j=n$ then there exists a
latent-class-specific set of parameters $\eta_{mn}$ which parametrizes
$f$, as well as a set of global hyper parameters $\theta$.

@@

\begin{equation}
  p(\vec{c}, \eta, \theta | R ) \propto \prod_{i, j} p(R(e_i, e_j) | f(d(e_i, e_j) | \eta_{c_ic_j}), \theta) \prod_{m, n} p(\eta_{mn} | \theta)  p(\theta) p(\vec{c} | \alpha) p(\alpha) p(\theta)
\end{equation}

@@

Our model is a extension of infinite stochastic block models
\autocite{Kemp2006a, other guys} to incorporate spatial constraints.

\todo{infinite subset relaitonal model has good reults}

@@

The ``logistic-distance Bernoulli'' model associates with each latent class 
two parameters, $\mu_{mn}$ and $\lambda_{mn}$, and the probability that two
cells $e_i$ and $e_j$ are connected is given by
\begin{eqnarray}
p^* &=& \frac{1.0}{1 + \exp \frac{d(e_i, e_j) - \mu_{mn}}{\lambda_{mn}}}\\
p &= & p^* \cdot (p_{max} - p_{min}) + p_{min}
\end{eqnarray}
where $p_{max}$ and $p_{min}$ are per-graph hyperparameters. Per-component parameters $\mu_{mn} \sim \exp(\mu | \mu^{hp})$ and $\lambda_{mn} \sim \exp(\lambda | \lambda^{hp})$. 
Inference is performed on these hyperparameters as well, see TODO supplemental 
for details.

@@

We then employ a transition kernel to update the per-component
parameter values $\eta_{mn}$. Conditioned on the assignment vector
$\vec{c}$ and the model hyperparameters $\theta, \alpha$ the 
individual $\eta_{mn}$ are indepednent. We slice sample \autocite{Neal2003}
each component's parameters, choosing the slice width as a function
of the global hyperparameter range.

@@

We obtained the connectome of c. elegans from \autocite, and isolated
the 279 nonpharyngeal neurons, with a total of 6393 chemical synapses
and 890 gap junctions originally cleaned up in \autocite{Chen2006}. A
cell's position was its distance along the anterior-posterior axis
normalzed between zero and one. We used both networks, the chemical
network as a directed graph and the electrical network as undirected
graph. We use the synapse counts with the logistic-distance poisson
likelihood, scaling the counts by a factor of 4 to compensate for 
the Poisson's overdispersion.

\subsection{Microprocessor}
We extracted the connection graph for all transistors in the MOS6502
\autocite{visual6502}. Each transistor has three terminals: an input
G, a source, and a drain, labeled C1 and C2. We identified the region
consisting of the three general-purpose registers X, Y, and S via
visual inspection and focused our efforts there. We created a total of
six connectivity matrices by examining possible terminal parings. One
graph, for example, $R^{gc_1}(e_i, e_j)=1$ if transistor $e_j$ and $e_j$ are
connected via pins $g$ and $c_1$. 
\todo{How do we handle symmetry here?}

% \begin{addendum}
%  \item Put acknowledgements here.
%  \item[Competing Interests] The authors declare that they have no
% competing financial interests.
%  \item[Correspondence] Correspondence and requests for materials
% should be addressed to A.B.C.~(email: myaddress@nowhere.edu).
% \end{addendum}
\section{Supplemental}
\subsection{Source code}
\todo{Put up website}

All source code and materials for running experiments can be
obtained from the project website, at http://spatialrelation.github.io.

\subsection{Extension to multiple graphs}
\todo{derivation of extension to multiple graphs}

\subsection{Hyperprior grids}
\todo{Plot hyperparameter Graphs}

\todo{Explain Adjusted Rand Index}
